SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-09-28 seqbias-supplement.tex
additions to the response letter
significance tests, minor tweaks
description of qpcr analysis
moste of r's notes on kl divergence

\title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
\author{Daniel C. Jones}

@@

\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
Sequencing Data}}{Daniel C. Jones}

@@

@@

>>>>

\title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
\author{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}

@@

\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
Sequencing Data}}{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}

@@

\section{Sequence bias in data from Au, et al., 2010}

\centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}

@@

\section{KL Divergence}

\subsection{TODO: Title}

Suppose $P$ and $Q$ are probability distributions defined on a discrete sample
space. E.g., in the context of this paper, think of them as the probabilities of
DNA $k$-mers for some fixed $k$. The \emph{Kullback-Leibler divergence}, also
known as \emph{relative entropy}, of $Q$ with respect to $P$ is defined as
$$H(Q||P) = \sum_{i} q_i \ln \frac{q_{i}}{p_{i}}$$
where $q_{i}$ ($p_{i}$) is the probability of observing the i$^{\text{th}}$
event according to the distribution $Q$ (resp., $P$), and the summation is taken
over all events in the sample space (e.g., all $k$-mers). In some sense, this is
a measure of the dissimilarity between the distributions: if $p_{i} \approx
q_{i}$ everywhere, their log rations will be near zero and $H$ will be small; as
$q_{i}$ and $p_{i}$ diverge, their log rotions will deviate from zero and $H$
will increase.

For a more quantitative and (perhaps) intuitive interpretation, consider the
following hypothesis-testing scenario. Given $m$ independent samples from either
$P$ (arbitrarily called the ``null'') or from $Q$ (the ``alternative''), how
large should $m$ be to cenfidently choose between these cases? A natural
approach is to use a likelihood ratio test: letting $Y_i$ be the number of times
event $i$ is observed in $m$ trials ($\sum Y_i = m$), the likelihood of the
sequence of observations under model $P$ is $$\prod_{i} p_{i}^{Y_{i}}$$
and similarily for $Q$. So, the logarithm of the likelihood ratios is
$$
\text{LLR} =
\ln \frac{\prod_{i} q_{i} ^ {Y_{i}}}{\prod_{i} p_{i} ^ {Y_{i}}}
= \sum_{i} Y_{i} \ln \frac{q_{i}}{p_{i}}
= m \sum_{i} \frac{Y_{i}}{m} \ln \frac{q_{i}}{p_{i}} $$

If the sample is drawn from the alternative distribution $Q$, then the
expectation of $frac{Y_{i}}{m}$ is $q_i$, so the expected value of the LLR is
$$ m \sum_{i} q_{i} \ln \frac{q_{i}}{p_{i}} = m H(Q || P) $$
That is, the KL divergence is exactly the expected per-sample contribution to
the log-likelihood ratio.
%% when comparing these two hypotheses ??
So, assuming the null hypothesis is false, in order for it to be rejected with
$1000:1$ odds, one should choose $m$ to be inversely proportional to $H(Q ||
P)$:
\begin{align*}
m H(Q||P) \ge \ln 1000 \\
m \ge \frac{\ln 1000}{H(Q || P)}
\end{align*}

As a concrete example, all of the RNA-Seq data sets examined in this paper show
a KL divergence between the 1$^{\text{st}}$ positon of reads versus the
transcriptomic background nucleotide distribution of $0.05$ or greater. Thus,
one can confidently reject the hypothesis that reads are beeing uniformly
sampled across the transcriptome by examining only a few hundred randomly
selected reads (i.e., $\ln 1000 / 0.05 \approx 140$).

\subsection{Accuracy of Multinomial Parameter Estimation}

Continuing the notation above, suppose $P$ as an unknown distribution with
parameters $p_1, \dots, p_r$, $\sum p_i = 1$ where $r$ is the number of points
in the sample space (e.g. $r = 4^{k}$ in the case of $k$-mers). Given a random
ample $X_1, X_2, \dots, X_r$ of size $n = \sum_{i} X_i$ from $P$, it is well
known that the maximum likelihood estimators for the parameters are $q_i =
\frac{X_i}{n} \approx p_i$. How good an estimate for $P$ is this distribution
$Q$?

$$E[q_i] = E\left[\frac{X_i}{n}\right] = \frac{E[X_i]}{n} = \frac{np_i}{n} =
p_i$$
And the standard deviation of each estimate is proportional to $1/\sqrt{n}$. So
these estimates are increasingly accurate as the sample size increases. A more
quantitative assessment of the accuracy of the estimator is obtained by
evaluating the KL divergence:
$$H(Q||P)
= \sum_{i = 1}^{r} q_i \ln \frac{q_i}{p_i}
= \sum_{i = 1}^{r} q_i \ln \left(1 + \frac{q_i - p_i}{p_i} \right) $$
Using the first two terms of he Taylor series for $\ln (1 + x)$, this is
\begin{align*}
H(Q||P) &\approx \sum_{i = 1}^{r} q_i \left( \frac{q_i - p_i}{p_i} -
\frac{1}{2} \left( \frac{q_i - p_i}{p_i} \right)^2 \right ) \\
&= \sum_{i = 1}^{r} q_i \frac{q_i - p_i}{p_i} -
\frac{q_i}{2 p_i} \frac{(q_i - p_i)^2} {p_i}
\end{align*}
Since $\sum_{i = 1}^{r} q_i = \sum_{i = 1}^{r} pi = 1$,
$\sum_{i = 1}^{r} p_i \frac{q_i - p_i}{p_i} = 0$, so
\begin{align*}
H(Q||P) &\approx \sum_{i=1}^{r} q_i \frac{q_i - p_i}{p_i} - p_i \frac{q_i -
p_i}{p_i} - \frac{q_i}{2 p_i} \frac{(q_i - p_i)^2}{p_i} \\
&= \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}\left(1 - \frac{q_i}{2 p_i}\right) \\
&\approx \frac{1}{2} \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}
\end{align*}
since $q_i \approx p_i$. Multiplying by $n^2 / n^2$ we have,
\begin{align*}
H(Q||P) &\approx \frac{1}{2n} \sum_{i=1}^{r} \frac{n q_i - n p_i)^2}{n p_i} \\
        &= \frac{1}{2n} \sum_{i=1}^{r} \frac{(X_i - E[X_i])^2}{E[X_i]}
\end{align*}

The summation is the test statistic for the $\chi^2$ goodness-of-fit test for a
multinomial distribution, and as $n \rightarrow \infty$ is known to follow a
$\chi^2$ distribution with $r - 1$ degrees of freedom. Finally, the expected
value of such a random variable is $r - 1$, hence the expected KL divergence of
the MLE inferred distribution $Q$ with respect to the true distribution $P$ is
$$E[H(Q||P)] = \frac{r - 1}{2n}$$

Stochastic simulations of this proves when $P$ is uniform show that the above
formula is a very good fit when $n > r$.
%% TODO: see figure X
